Learning and comparing functional connectomes across subjects

### 2013-02-21 paper.tex
iter

% General introduction
Functional connectivity reveals the synchronization of distant neural systems
via correlations in neurophysiological measures of brain activity 
\cite{friston1993, biswal1995}.  Given that
high-level function emerges from the interaction of specialized units
\cite{tononi1992},
functional connectivity is an essential part of the description of brain
function, that complements the localizationist picture emerging from the
systematic mapping of regions recruited in tasks \cite{sporns2004}. However,
while there exists a well-defined standard analysis framework for activation
mapping that enables statistically-controlled comparisons across subjects
\cite{friston1995}, group-level analyses of functional connectivity still face
many open methodological challenges. Deriving a picture of a single subject's
functional connectivity is by itself not straightforward, as the brain comprises
a myriad of interacting subsystems and its connectivity must be decomposed into
simplified and synthetic representations. An important view of brain
connectivity is that of distributed functional networks depicted by their
spatial maps \cite{fox2005}. Another no less important and complementary view is
that of connections linking localized functional modules depicted as a graph
\cite{bullmore2009}. This representation of brain connectivity is often called
the functional connectome \cite{sporns2005} and is the focus of intense
worldwide research efforts as it holds promises of new insights in cognition and
pathologies \cite{greicius2008b,biswal2010,fox2010}.

@@

In addition to standard preprocessing performed for task based analysis
(slice-timing correction, realignment, spatial normalization, and possibly
smoothing), connectivity based analysis require additional denoising 
to separate intrinsic activity from confounding signals. This process involves
regressing time series capturing sources of structured noise from the fMRI data.
Physiological noise due to cardiac and respiration are two important
noise signals \cite{hu1995, lund2006, birn2006, birn2008} that are
difficult to control for and as a result are not commonly regressed out. Instead
the mean signal from white matter (WM) and cerebrospinal fluid (CSF) are used as
surrogates to measure these sources of noise as well as other scanner induced
signal fluctuations \cite{fox2005, lund2006}. More complex models account for
spatial variation in noise by incorporating voxel specific regressors of
neighboring WM (ANATICOR \cite{jo2010}) or the top components from a
principal
components analysis of WM signal (CompCor \cite{behzadi2007}). Head motion
induced signal fluctuations are accounted for by incorporating movement
parameters \cite{friston1996, fox2005, lund2006}. The global mean time series
has been proposed as an additional noise regressor that appears to improve the
spatial specificity of connectivity results \cite{fox2005, fox2009}. This practice
has become controversial since the global signal regression introduces negative correlations
\cite{murphy2009,chang2009,saad2012}. 
Removing these sources of
nuisance in addition to linear trends results in more contrasted
correlation matrices that improve the delineation of functional structures
(fig.\,\ref{fig:correlation_matrices}).

@@

\paragraph{Graph-topological metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are grouped under the terms of ``graph theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they have a relatively
small number of connections, any two regions of the brain are well
connected. 
Another interesting consequence of their specific topology is
the resilience it gives the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes, the local clustering
coefficients, or the node degree centrality \cite{rubinov2010}. Given that pathologies without a
localized focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, the graph-topological metrics are
promising markers to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that these properties are not
specific to brain function: correlation matrices display
small-word properties such as local clustering by construction. Indeed, if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlights
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices, as discuss in the previous section.

@@

Predictive modeling is concerned with learning (or fitting) a model that is
capable of \emph{predicting} information from unseen data
\cite{pereira2009}.  In the context of connectomes, predictive modeling
can extract connectivity-based biomarkers of disease diagnosis,
prognosis, or other phenotypic outcomes \cite{craddock2009, dosenbach2010}. The
accuracy of a predictive model provides a measure of the amount of information
present in the connectome about the phenotypic measure being evaluated
\cite{kjems2002, kriegeskorte2006}.  When combined with reproducibility, prediction accuracy
provides a metric for evaluating experimental trade-offs for data acquisition,
preprocessing, and analysis \cite{strother2002, laconte2003}.  Multivariate
predictive models are attractive in connectomics because they are sensitive to dependencies
between features and avoid the need to correct for multiple comparisons since
the significance of an entire pattern is evaluated using a single statistical
test. 
Additionally modern predictive modeling techniques draw from the
statistical learning literature, which specifically addresses
 high dimensional datasets with few observations.
Predictive modeling has been
successfully applied to identify connectome-based biomarkers of Alzheimer's
disease \cite{stonnington2010}, depression \cite{craddock2009, zeng2012}, schizophrenia
\cite{cecchi2009, shen2010}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}. The growing interest for applying
predictive modeling to connectivity analysis was highlighted by the ADHD200
Global Competition, in which the object was to identify a connectivity based
biomarker of ADHD \cite{adhd2002012}. Recent work has illustrated the utility
of predictive modeling for deriving connectivity models at the individual level
\cite{chu2011}.

% A bit of technical details
% technical details
Technically, predictive modeling is a supervised machine learning problem where
a target to be predicted --\emph{e.g.}\ age, disease state, cognitive state-- is
available for each observation of the data. In the context of comparing
connectomes, features used in the predictive model correspond to bivariate
measures of connectivity, or any of the previously discussed graph summary metrics.  
%
The quality of a predictive model is determined by its prediction accuracy (or
generalization ability) which is measured using one or more iterations of
cross-validation. Cross-validation iteratively subdivides available data into a
subset used for training the classifier and a dataset for evaluating classifier
performance\footnote{Several strategies exist for performing cross-validation and the commonly
used approach of using only a single observation for testing (leave-one-out
cross-validation) results in highly variable estimates of prediction accuracy
\cite{friedman2001}.
Alternative approaches such as (5 or 10)-fold cross-validation,
or $0.632+$ bootstrap should be preferred \cite{friedman2001}.} 
 \cite{pereira2009}.
The significance of
achieved prediction accuracy can be assessed using permutation tests
\cite{golland2003}.
Predictive modeling approaches typically require the specification of several
parameters, which may be chosen based on domain specific knowledge or
requirements \cite{cherkassky1998}, determined using an analytical approach
\cite{cherkassky2004}, or optimized using a second level cross validation procedure
\cite{friedman2001}.

>>>>

% General introduction
Functional connectivity reveals the synchronization of distant neural systems
via correlations in neurophysiological measures of brain activity 
\cite{friston1993, biswal1995}.  Given that
high-level function emerges from the interaction of specialized units
\cite{tononi1992},
functional connectivity is an essential part of the description of brain
function, that complements the localizationist picture emerging from the
systematic mapping of regions recruited in tasks \cite{sporns2004}. However,
while there exists a well-defined standard analysis framework for activation
mapping that enables statistically-controlled comparisons across subjects
\cite{friston1995}, group-level analysis of functional connectivity still face
many open methodological challenges. Deriving a picture of a single subject's
functional connectivity is by itself not straightforward, as the brain comprises
a myriad of interacting subsystems and its connectivity must be decomposed into
simplified and synthetic representations. An important view of brain
connectivity is that of distributed functional networks depicted by their
spatial maps \cite{fox2005}. Another no less important and complementary view is
that of connections linking localized functional modules depicted as a graph
\cite{bullmore2009}. This representation of brain connectivity is often called
the functional connectome \cite{sporns2005} and is the focus of intense
worldwide research efforts as it holds promises of new insights in cognition and
pathologies \cite{greicius2008b,biswal2010,fox2010}.

@@

In addition to standard preprocessing performed for task based analysis
(slice-timing correction, realignment, spatial normalization, and possibly
smoothing), connectivity based analysis require additional denoising 
to separate intrinsic activity from confounding signals. This process involves
regressing time series capturing sources of structured noise from the fMRI data.
Physiological noise due to cardiac and respiration are two important
noise signals \cite{hu1995, lund2006, birn2006, birn2008} that are
difficult to control for and as a result are not commonly regressed out. Instead
the mean signal from white matter (WM) and cerebrospinal fluid (CSF) are used as
surrogates to measure these sources of noise as well as other scanner induced
signal fluctuations \cite{fox2005, lund2006}. More complex models account for
spatial variation in noise by incorporating voxel-specific regressors of
neighboring WM (ANATICOR \cite{jo2010}) or the top components from a
principal
components analysis of WM signal (CompCor \cite{behzadi2007}). Head motion
induced signal fluctuations are accounted for by incorporating movement
parameters \cite{friston1996, fox2005, lund2006}. The global mean time series
has been proposed as an additional noise regressor that appears to improve the
spatial specificity of connectivity results \cite{fox2005, fox2009}. This practice
has become controversial since the global signal regression introduces negative correlations
\cite{murphy2009,chang2009,saad2012}. 
Removing these sources of
nuisance in addition to linear trends results in more contrasted
correlation matrices that improve the delineation of functional structures
(fig.\,\ref{fig:correlation_matrices}).

@@

\paragraph{Graph-topological metrics}
%
Functional connectivity graphs have been found to display specific
topological properties\footnote{In the neuroscience world, these
descriptions are grouped under the terms of ``graph-theoretical
approaches'', however graph theory is an entire division of mathematics
and computer science that is concerned with much more than topology of
random graphs.} that are characteristic of small-word networks
\cite{stam2004,salvador2005,achard2006,bullmore2009}. These networks
display excellent transport properties: although they have a relatively
small number of connections, any two regions of the brain are well
connected. 
Another interesting consequence of their specific topology is
the resilience it gives the system to attacks such as resulting from
brain lesions \cite{achard2006}. This overall structure of
functional-connectivity graphs can be summarized by a few metrics, such
as the average path length between any two nodes, the local clustering
coefficients, or the node degree centrality \cite{rubinov2010}. Given that pathologies without a
localized focus, such
as schizophrenia, are thought to have a global impact on brain
connectivity \cite{liu2008,bassett2008}, the graph-topological metrics are
promising markers to perform inter-subject comparison. Such an
approach is appealing as it is not subject to multiple comparison issues.
However, it has been criticized as giving a fairly unspecific
characterization of the brain and being fragile to noise
\cite{ioannides2007}. Another caveat is that these properties are not
specific to brain function: correlation matrices display
small-word properties such as local clustering by construction. Indeed, if two
nodes are strongly correlated to a third, they are highly likely to be
correlated to each other \cite{zalesky2012}. This observation highlights
the need for well defined null-hypothesis \cite{zalesky2012,rubinov2011},
but also for controlled recovery of brain functional connectivity going
beyond empirical correlation matrices, as discuss in the previous section.

@@

Predictive modeling is concerned with learning (or fitting) a model that is
capable of \emph{predicting} information from unseen data
\cite{pereira2009}.  In the context of connectomes, predictive modeling
can extract connectivity-based biomarkers of disease diagnosis,
prognosis, or other phenotypic outcomes \cite{craddock2009, dosenbach2010}. The
accuracy of a predictive model provides a measure of the amount of information
present in the connectome about the phenotypic measure being evaluated
\cite{kjems2002, kriegeskorte2006}.  When combined with reproducibility, prediction accuracy
provides a metric for evaluating experimental trade-offs for data acquisition,
preprocessing, and analysis \cite{strother2002, laconte2003}.  Multivariate
predictive models are attractive in connectomics because they are sensitive to dependencies
between features and avoid the need to correct for multiple comparisons since
the significance of an entire pattern is evaluated using a single statistical
test. 
Additionally modern predictive modeling techniques draw from the
statistical learning literature, which specifically addresses
 high dimensional datasets with few observations.
Predictive modeling has been
successfully applied to identify connectome-based biomarkers of Alzheimer's
disease \cite{stonnington2010}, depression \cite{craddock2009, zeng2012}, schizophrenia
\cite{cecchi2009, shen2010}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}. The growing interest for applying
predictive modeling to connectivity analysis was highlighted by the ADHD200
Global Competition, in which the object was to identify a connectivity-based
biomarker of ADHD \cite{adhd2002012}. Recent work has illustrated the utility
of predictive modeling for deriving connectivity models at the individual level
\cite{chu2011}.

% A bit of technical details
% technical details
Technically, predictive modeling is a supervised machine learning problem where
a target to be predicted --\emph{e.g.}\ age, disease state, cognitive state-- is
available for each observation of the data. In the context of comparing
connectomes, features used in the predictive model correspond to bivariate
measures of connectivity, or any of the previously discussed graph summary metrics.  
%
The quality of a predictive model is determined by its prediction accuracy (or
generalization ability) which is measured using one or more iterations of
cross-validation. Cross-validation iteratively subdivides available data into a
subset used for training the classifier and a dataset for evaluating classifier
performance\footnote{Several strategies exist for performing cross-validation and the commonly
used approach of using only a single observation for testing (leave-one-out
cross-validation) results in highly variable estimates of prediction accuracy
\cite{friedman2001}.
Alternative approaches such as (5 or 10)-fold cross-validation,
or $0.632+$ bootstrap should be preferred \cite{friedman2001}.} 
 \cite{pereira2009}.
The significance of
achieved prediction accuracy can be assessed using permutation tests
\cite{golland2003}.
Predictive modeling approaches typically require the specification of several
parameters, which may be chosen based on domain specific knowledge or
requirements \cite{cherkassky1998}, determined using an analytical approach
\cite{cherkassky2004}, or optimized using a second-level cross validation procedure
\cite{friedman2001}.
